Comparing probabilistic predictive models applied to football

### 2015-07-08 response.tex

\documentclass[11pt]{article}
\usepackage[latin1]{inputenc}
\usepackage{amsmath, graphicx, layout, verbatim}
\usepackage{setspace}
\usepackage{fancyhdr}
\usepackage{float}
\usepackage{color}
\usepackage{natbib}
\usepackage{verbatim}
\usepackage{subfig}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[pagebackref,colorlinks=true,urlcolor=blue,citecolor=blue	,linkcolor=blue]{hyperref}
\textheight25cm \footskip0.5cm \topmargin-1.5cm \headsep0.2cm
\oddsidemargin-0.5cm \evensidemargin-0.5cm
 \marginparwidth0cm \marginparsep = 0pt
\textwidth16cm

@@

\vspace{2mm}
\textbf{Response:} We thank the referee for taking the time reviewing the paper. Below we list our detailed response to the referee's questions.\\

\vspace{6mm}

\hspace{4mm} \textbf{Reviewer:} \textit{\textbf{General Comments.}
	What the authors have shown in my opinion is that the blackbox predictors they consider as
	competitors do not incorporate covariate information effectively and/or the covariates are not
	informative in learning about game outcomes. Even though it is not the focus of the paper,
	it seems to me that including some detail (at least on an intuitive level) of how information is
	used in these black box methods would be helpful in better understanding why they perform so
	poorly relative to the very unsophisticated MD models.}

\vspace{2mm}
\textbf{Response: we added subsection in Section 2 explaining some details of the benchmark models.***Falta o Ernesto***.}

### 2015-07-08 response.tex

\vspace{2mm}
\textbf{Response:} We thank the referee for taking the time reviewing the paper. Below we list our detailed response to the referee's questions.\\

\vspace{6mm}

\hspace{4mm} \textbf{Reviewer:} \textit{\textbf{General Comments.}
	What the authors have shown in my opinion is that the blackbox predictors they consider as
	competitors do not incorporate covariate information effectively and/or the covariates are not
	informative in learning about game outcomes. Even though it is not the focus of the paper,
	it seems to me that including some detail (at least on an intuitive level) of how information is
	used in these black box methods would be helpful in better understanding why they perform so
	poorly relative to the very unsophisticated MD models.}

\vspace{2mm}
\textbf{Response: We included a new subsection
	(Subsection 2.1) in Section 2 giving more details about the
	benchmark models considered, which are similar to many other
	approaches in the literature. In fact, we believe that football
	prediction is a hard task. This is mainly because several
	factors may influence the final outcome of a match, such
	as referee's mistakes, expulsions or injuries of important
	players, weather conditions, among many others. All those factors are difficult
	(maybe impossible) to take into account in any statistical model.
	Thus, the use of elaborate models leads to small gain in prediction
	accuracy when compared to simpler models, because of the great
	unpredictability of football matches.}

@@

\vspace{2mm}
\textbf{Response: we changed ``previsions'' by ``predictions''. There was a mistake on p. 4 because we should refer to the ``second round''. We fixed this and added a noted explaining that each team plays 19 matches each round, since the championship has 20 teams. More details on the championship rule are given in Section 4 (Results).}

\vspace{6mm}

\hspace{4mm} \textbf{Reviewer:} \textit{\textbf{Section 2: The models: theoretical background.} 
	\begin{itemize}
		\item You point out that the assumptions associated for the MD models and black-box models
		are different. Does this make comparing them untenable? If not, why?
	\end{itemize}}

	\vspace{2mm}
	\textbf{Response: We believe that iwhen the model assumptions are different, the only possible comparison are in terms
		of predictive accuracy. In the present study, this comparison is possible since all the models report the probabilities of win, draw or loss of the home team.
		Furthermore, from the perspective of football fans the most important feature of a model is to provide ``good'' predictions.}

### 2015-08-14 template.tex

    The famous saying of John von Neumann: {\it ``With four parameters I can fit an elephant, and with five I can make him wiggle his trunk''} is a perhaps just a humorous version of Ockham's razor:
    {\it``Entities are not to be multiplied without necessity''}, but illustrates very well a principle that is not often followed by researchers, especially when building predictive models.
    This is true also in the literature of probabilistic sports predictions, particularly regarding models devoted to football or soccer (henceforth, football) predictions.

    Bothered by such concerns, our goal was to answer the following question: is it possible, using a simple probabilistic model, to predict results of football matches better than more sophisticated or complex models?
    To answer this, we compared two sophisticated models that consider multiple covariates with simple multinomial-Dirichlet models that consider only the number of matches won, tied or lost by each team as inputs.
    The models were tested for the second round of the Brazilian football championships of first division and compared using standard metrics or scoring rules.
    We also used other criteria such as the proportion of matches that were predicted ``correctly'' by each model and a measure of calibration, both explained in detail below.
    According to the adopted criteria, the simple models showed a similar predictive performance to the complex ones.

    The predictions of the two benchmark models were published on Internet websites before every matchday, and are treated by us as black-box models (BB1 and BB2).
    Since one of the authors is responsible for one of these models and the other was described in a dissertation \citep{arruda2000}, we know that both models are based on Holgate bivariate Poisson regression considering as explanatory variables the home field advantage, attack and defense strength for each team, and provide the probabilities of all possible outcomes (home team wins, draws or loses) for each match.
    Because both models are based on elaborate parametric assumptions, they are suitable for the comparison we are proposing: simple versus complex models. To be precise, we would have to define what is a ``simple'' model and what is a ``complex'' model. Based on the opening quotations, one may say that the number of parameters defines the complexity of a model, which is not exactly the definition we will use, since we consider other criteria, like the interpretation of explanatory variables and mathematical structure of the statistical model.

@@

    This paper is organized as follows.
    Section \ref{sec::experimental} describes the proposed models, Section \ref{sec::scoring} presents the scoring rules and criteria used in this work to classify the models.
    Section \ref{sec::results} reports the performance of the models predicting 190 matches comprising the second round of the Brazilian championship of first division--each team plays 19 matches each round, since the championship has 20 teams.
    In Section \ref{sec::remarks} we discuss the results and close with proposals for future research.

@@

    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \subsection{Benchmark Models}
    \label{sec::Benchmark}

    The benchmark models assume that the number of goals $(Y_1, Y2)$ scored respectively by teams A (home team) and B (away
    team) follows a bivariate Poisson distribution \citep{Holgate64} with parameters $(\lambda_1, \lambda_2, \lambda_3)$,
    which has p.m.f given by
\begin{equation*}
P(Y_1 = y_1, Y_2 = y_2 | \lambda_1, \lambda_2, \lambda_3) =
\exp\{-(\lambda_1 + \lambda_2 + \lambda_3)\}
 \sum_{k = 0}^{\min(y_1, y_2)} \dfrac{\lambda_1^{x - k} \lambda_2^{y - k} \lambda_3^k}{(x-k)!(y -
 k)!k!}, \label{eq::pois.biv}
\end{equation*}
for $\lambda_1, \lambda_2 > 0$ and $\lambda_3 \geq 0$.

Both marginal distributions of $(Y_1, Y_2)$ follow Poisson
distributions with dependence parameter $\lambda_3 \geq 0$. If
$\lambda_3 = 0$ the marginal distributions are independent, while if
$\lambda_3 > 0$ the marginal distributions are positively
correlated. Because of its flexibility, \cite{Karlis2003} argues
that this distribution is a plausible choice for modeling dependence
of scores in sports competitions.

@@

Similar to \cite{Lee97} and \cite{Karlis2003}, both benchmark models
adopt the following log-linear link functions
\begin{align*}
\log(\lambda_1) &= \mu + \text{ATT}_A - \text{DEF}_B + \gamma, \\
\log(\lambda_2) &= \mu + \text{ATT}_B - \text{DEF}_A,
\end{align*}
where $\mu$ is a parameter representing the average number of goals
in a match, $\text{ATT}_k$ is the offensive strength of team k,
$\text{DEF}_k$ is the defensive strength of team k and $\gamma$ is
the home advantage for team A, $k = A, B$.

The prediction for an upcoming match-day is obtained by fitting the
model to all the observed scores up to the previous match-day and
then summing up the probabilities of all scores relevant to the win,
draw and loss outcomes of future matches.

@@

        In this work, we are interested in checking the predictive performance of the Multinomial-Dirichlet
        model for the result of every single match in the second round of the championship given the results
        of all the previous matches observed in the same tournament. Predictions
        are carried out by using a Bayes information updating mechanism. First, we use full-time
        results from the first round as historical data for construction of the Dirichlet prior: we assign a uniform prior on
        $\boldsymbol{\theta}$ over the 2-simplex, i.e. $\mathcal{D}(1, 1, 1)$, and combine this prior with
        the first-round data, obtaining a posterior Dirichlet distribution through conjugacy that represents
        the information about $\boldsymbol{\theta}$ up to the first round. Then, the posterior of the first
        round becomes the prior for the second round, which, for every matchday in the second round, will be combined with all the observed second-round matches
        up to that matchday in order to yield posterior predictive distributions. Note that, on the other hand, the two
        black-box models make predictions from a frequentist perspective; BB1 uses scores of every match
        in the previous two years, while BB2 uses the match scores of the current championship only.

        Before we explain and illustrate the proposed models with an example, we make two further remarks.
        The first one is that we will separately consider home and away games for each team. This will allow us to take into account the different performances under those conditions.
        The second remark is that the prediction of an upcoming match between two teams $A$ and $B$ can be done using
        the past performance of either team $A$ or $B$.
        To solve this problem, a compromise between the two predictive distributions was adopted.
        The next subsections \ref{sec::Mn_Dir1}, \ref{sec::Mn_Dir2} and \ref{sec::Mn_Dir3} present three different possible solutions.

@@

    \section{Scoring rules and calibration}
    \label{sec::scoring}

@@

    Formally, let $X$ be a random variable
    taking values in $\mathcal{X}=\{1,2,3\}$ indicating
    the outcome of the match, with 1 standing for home win, 2 for draw and 3 for away win.  Moreover, let $P=(P_1,P_2,P_3)$ denote one's probabilistic prediction
    about $X$, i.e., $P$ lies in the 2-simplex set $\Delta_2=\{(p_1,p_2,p_3):p_1+p_2+p_3=1, \ p_1,p_2,p_3\geq0\}$ (see Figure
    \ref{fig:simplex}).
    A scoring rule is a function
    that assigns a real number (score) $S(x,P)$ to each $x \in \mathcal{X}$
    and $P \in \Delta_2$
    such that
    for any given $x$ in $\mathcal{X}$, the score  $S(x,P)$ is minimized when $P$ is
    $(1,0,0)$, $(0,1,0)$ or $(0,0,1)$ depending if $x$ is 1, 2 or 3, respectively.
    The score $S(x,P)$ can be thought as
    a penalty to be paid when one assigns the
    probabilistic prediction $P$ and outcome
    $x$ occurs. Also, the ``best'' possible score (i.e., the smallest score value) is achieved when the probabilistic prediction for the outcome of the game is perfect. A scoring rule may also be defined to be such that a large value of the score indicates good forecasts.
        %Most of the scoring rules also exhibit an arbitrary normalizing feature such as $S(x,P_X)=0$ for each $x$.

@@

    Say a model's probabilistic prediction for $X$ is $P^*=(P_1^*,P_2^*,P_3^*).$ A proper scoring rule $S$ is a scoring rule  such that the mean score value
    $$E_{P^*}[S(X,P)]=\sum_{x=1}^3 S(x,P)P^*_x$$
     is minimized when $P=P^*$.
    In other words, if one announces
    $P$ as his probabilistic prediction
    and uses $S$ as scoring rule, the lowest
    expected penalty is obtained by reporting the correct model's  uncertainty about $X$, $P^*$.
    Thus, the use of a proper scoring rule encourages the forecaster to announce $P^*$ (the correct one)
    as his probabilistic prediction  rather than some other quantity.
    %Moreover, this method automatically leads to an overall comparison between the outcomes of different personal evaluations concerned with the same totality of events and, therefore, the accumulated score may be considered a concrete measure of success of each model.

@@

    Figure \ref{fig::entropy} presents the boxplots of the entropy of each prediction for the predictive models considered. Recall
        that the entropy of a prediction $(p_1,p_2,p_3)$ is given by $- \sum_{i=1}^3 p_i \log{p_i}$.
    Since the entropy values are smaller than 1.09, the plots show that the predictions
    are typically more informative than the trivial prediction. Nevertheless, all methods yielded similar entropies in average, that is,
    none of them seem to provide more informative probabilities.

@@

    Although our benchmark models were here defined as complex, we can find much more elaborate models in the literature, hindering the estimation and prediction processes.
    Among them, we can cite those based on modeling the match as a stochastic process evolving through time \citep{Dixon98, Volf2009, Titman2015}, those allowing for the team performance parameters to change along the season \citep{Rue2000, Crowder2002, Owen2011, Koopman2015} and those modeling dependence between number of goals by means of bivariate count distributions \citep{Dixon97, Karlis2003, McHale2007, McHale2011}. Models that predict the number of goals have more specific objectives than those prediciting the final outcomes and therefore require more elaborate modeling, justifying the complexity of such models.

@@

    This is often unnoticed by users of such methods, which leads to the overuse of models that are
    hard to interpret and to implement, and that also require one to keep track of several covariates. We would call this phenomenon the {\it king's nudity} in allusion to Anderson's {\it The Emperor's New Clothes}, \cite{emperor}: ``But he isn't wearing anything at all!''. However, this is not always the case. In this spirit, since we have explored only the prediction of football matches, we would like to study if the same extends to other fields such as, for instance, weather or economics forecasts.

>>>>

    The famous saying of John von Neumann: {\it ``With four parameters I can fit an elephant, and with five I can make him wiggle his trunk''} is a perhaps just a humorous version of Ockham's razor:
    {\it``Entities are not to be multiplied without necessity''}, but illustrates very well a principle that is not often followed by researchers, especially when building predictive models.
    This is true also in the literature of probabilistic sports predictions, particularly regarding models devoted to football or soccer {\color{red}(henceforth, football)} predictions.

    Bothered by such concerns, our goal was to answer the following question: is it possible, using a simple probabilistic model, to predict results of football matches better than more sophisticated or complex models?
    To answer this, we compared two sophisticated models that consider multiple covariates with simple multinomial-Dirichlet models that consider only the number of matches won, tied or lost by each team as inputs.
    The models were tested for the second round of the Brazilian football championship of first division and compared using standard metrics or scoring rules.
    We also used other criteria such as the proportion of matches that were predicted ``correctly'' by each model and a measure of calibration, both explained in detail below.
    According to the adopted criteria, the simple models showed a similar predictive performance to the complex ones.

    The predictions of the two benchmark models were published on Internet websites before every matchday, and are treated by us as black-box models (BB1 and BB2).
    {\color{red} Since one of the authors is responsible for one of these models and the other was described in a dissertation \citep{arruda2000}}, we know that both models are based on Holgate bivariate Poisson regression considering as explanatory variables the home field advantage, attack and defense strength of each team, and provide the probabilities of all possible outcomes (home team wins, draws or loses) for each match.
    Because both models are based on elaborate parametric assumptions, they are suitable for the comparison we are proposing: simple versus complex models.

    {\color{red} To be precise, we would have to define what is a ``simple'' model and what is a ``complex'' model. Based on the opening quotations, one may say that the number of parameters defines the complexity of a model, which is not exactly the definition we will use, since we consider other criteria, like the interpretation of explanatory variables and mathematical structure of the statistical model.}

@@

    This paper is organized as follows.
    Section \ref{sec::experimental} describes the proposed models, Section \ref{sec::scoring} presents the scoring rules and criteria used in this work to classify the models.
    Section \ref{sec::results} reports the performance of the models predicting 190 matches comprising the second round of the Brazilian championship of first division---{\color{red}each team plays 19 matches each round, since the championship has 20 teams.}
    In Section \ref{sec::remarks} we discuss the results and close with proposals for future research.

@@

    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    {\color{red} \subsection{Benchmark Models}
    \label{sec::Benchmark}

    The benchmark models assume that the number of goals $(Y_1, Y2)$ scored respectively by teams $A$ (home team) and $B$ (away
    team) has a bivariate Poisson distribution \citep{Holgate64} with parameters $(\lambda_1, \lambda_2, \lambda_3)$,
    which has probability mass function given by
\begin{equation*}
P(Y_1 = y_1, Y_2 = y_2 | \lambda_1, \lambda_2, \lambda_3) =
\exp\{-(\lambda_1 + \lambda_2 + \lambda_3)\}
 \sum_{k = 0}^{\min(y_1, y_2)} \dfrac{\lambda_1^{x - k} \lambda_2^{y - k} \lambda_3^k}{(x-k)!(y -
 k)!k!}, \label{eq::pois.biv}
\end{equation*}
for $\lambda_1, \lambda_2 > 0$ and $\lambda_3 \geq 0$.

Both marginal distributions of $(Y_1, Y_2)$ has Poisson
distributions with dependence parameter $\lambda_3 \geq 0$. If
$\lambda_3 = 0$ the marginal distributions are independent, while if
$\lambda_3 > 0$ the marginal distributions are positively
correlated. Because of its flexibility, \cite{Karlis2003} argues
that this distribution is a plausible choice for modeling dependence
of scores in sports competitions.

@@

Similar to \cite{Lee97} and \cite{Karlis2003}, both benchmark models
adopt the following log-linear link functions
\begin{align*}
\log(\lambda_1) &= \mu + \text{ATT}_A - \text{DEF}_B + \gamma, \\
\log(\lambda_2) &= \mu + \text{ATT}_B - \text{DEF}_A,
\end{align*}
where $\mu$ is a parameter representing the average number of goals
in a match, $\text{ATT}_k$ is the offensive strength of team $k$,
$\text{DEF}_k$ is the defensive strength of team $k$ and $\gamma$ is
the home advantage for team $A$, $k = A, B$.

The prediction for an upcoming matchday is obtained by fitting the
model to all the observed scores up to the previous match-day and
then summing up the probabilities of all scores relevant to the win,
draw and loss outcomes of future matches.}

@@

        {\color{red} In this work, we are interested in checking the predictive performance of the Multinomial-Dirichlet
        model for the result of every single match in the second round of the championship given the results
        of all the previous matches observed in the same tournament. Predictions
        are carried out by using a Bayes information updating mechanism. First, we use full-time
        results from the first round as historical data for construction of the Dirichlet prior: we assign a uniform prior on
        $\boldsymbol{\theta}$ over the 2-simplex, i.e. $\mathcal{D}(1, 1, 1)$, and combine this prior with
        the first-round data, obtaining a posterior Dirichlet distribution through conjugacy that represents
        the information about $\boldsymbol{\theta}$ up to the first round. Then, the posterior of the first
        round becomes the prior for the second round, which, for every matchday in the second round, will be combined with all the observed second-round matches
        up to that matchday in order to yield posterior predictive distributions. Note that, on the other hand, the two
        black-box models make predictions from a frequentist perspective; BB1 uses scores of every match
        in the previous two years, while BB2 uses the match scores of the current championship only.}

        Before we explain and illustrate the proposed models with an example, we make two further remarks.
        The first one is that we will separately consider home and away games for each team, allowing us to take into account the different performances under these conditions.
        The second remark is that the prediction of an upcoming match between two teams $A$ and $B$ can be done using
        the past performance of either team $A$ or $B$.
        To solve this problem, a compromise between the two predictive distributions was adopted.
        The next subsections \ref{sec::Mn_Dir1}, \ref{sec::Mn_Dir2} and \ref{sec::Mn_Dir3} present three different possible solutions.

@@

    {\color{red}\section{Scoring rules and calibration}
    \label{sec::scoring}

@@

    Formally, let $X$ be a random variable
    taking values in $\mathcal{X}=\{1,2,3\}$ indicating
    the outcome of the match, with 1 standing for home win, 2 for draw and 3 for away win.  Moreover, let $P=(P_1,P_2,P_3)$ denote one's probabilistic prediction
    about $X$, i.e., $P$ lies in the 2-simplex set $\Delta_2=\{(p_1,p_2,p_3):p_1+p_2+p_3=1, \ p_1,p_2,p_3\geq0\}$ (see Figure
    \ref{fig:simplex}).
    A scoring rule is a function
    that assigns a real number (score) $S(x,P)$ to each $x \in \mathcal{X}$
    and $P \in \Delta_2$
    such that
    for any given $x$ in $\mathcal{X}$, the score  $S(x,P)$ is minimized when $P$ is
    $(1,0,0)$, $(0,1,0)$ or $(0,0,1)$ depending if $x$ is 1, 2 or 3, respectively.
    The score $S(x,P)$ can be thought as
    a penalty to be paid when one assigns the
    probabilistic prediction $P$ and outcome
    $x$ occurs. Also, the ``best'' possible score (i.e., the smallest score value) is achieved when the probabilistic prediction for the outcome of the game is perfect. A scoring rule may also be defined to be such that a large value of the score indicates good forecasts.

@@

    Say a model's probabilistic prediction for $X$ is $P^*=(P_1^*,P_2^*,P_3^*).$ A proper scoring rule $S$ is a scoring rule  such that the mean score value
    $$E_{P^*}[S(X,P)]=\sum_{x=1}^3 S(x,P)P^*_x$$
     is minimized when $P=P^*$.
    In other words, if one announces
    $P$ as his probabilistic prediction
    and uses $S$ as scoring rule, the lowest
    expected penalty is obtained by reporting the correct model's  uncertainty about $X$, $P^*$.
    Thus, the use of a proper scoring rule encourages the forecaster to announce $P^*$ (the correct one)
    as his probabilistic prediction  rather than some other quantity.}
    %Moreover, this method automatically leads to an overall comparison between the outcomes of different personal evaluations concerned with the same totality of events and, therefore, the accumulated score may be considered a concrete measure of success of each model.

@@

    Figure \ref{fig::entropy} presents the boxplots of the entropy of each prediction for the predictive models considered. {\color{red} Recall
        that the entropy of a prediction $(p_1,p_2,p_3)$ is given by $- \sum_{i=1}^3 p_i \log{p_i}$.}
    Since the entropy values are smaller than 1.09, the plots show that the predictions
    are typically more informative than the trivial prediction. Nevertheless, all methods yielded similar entropies in average, that is,
    none of them seem to provide more informative probabilities.

@@

    Although our benchmark models were here defined as complex, we can find much more elaborate models in the literature, hindering the estimation and prediction processes.
    Among them, we can cite those based on modeling the match as a stochastic process evolving through time \citep{Dixon98, Volf2009, Titman2015}, those allowing for the team performance parameters to change along the season \citep{Rue2000, Crowder2002, Owen2011, Koopman2015} and those modeling dependence between number of goals by means of bivariate count distributions \citep{Dixon97, Karlis2003, McHale2007, McHale2011}. {\color{red} Models that predict the number of goals have more specific objectives than those prediciting the final outcomes and therefore require more elaborate modeling, justifying the complexity of such models}.

@@

    This is often unnoticed by users of such methods, which leads to the overuse of models that are
    hard to interpret and to implement, and that also require one to keep track of several covariates. We would call this phenomenon the {\it king's nudity} {\color{red} in allusion to Anderson's {\it The Emperor's New Clothes}, \cite{emperor}: ``But he isn't wearing anything at all!''}. However, this is not always the case. In this spirit, since we have explored only the prediction of football matches, we would like to study if the same extends to other fields such as, for instance, weather or economics forecasts.
